---
title: "HR Data Harmonization"
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# # Required packages for the vignette
# cran_pkgs <- c("polyglotr", "DT")
# 
# # Install missing CRAN packages
# for (pkg in cran_pkgs) {
#   if (!requireNamespace(pkg, quietly = TRUE)) {
#     install.packages(pkg)
#   }
# }
# 
# # Install missing GitHub packages
# if (!requireNamespace("labourR", quietly = TRUE)) {
#   if (!requireNamespace("remotes", quietly = TRUE)) {
#     install.packages("remotes")
#   }
#   remotes::install_github("AleKoure/labourR")
# }


library(reactable)
library(govhr)
library(dplyr)
library(knitr)
library(data.table)
library(polyglotr)
library(labourR)
library(DT)
library(lubridate)

```

# Introduction

Across many countries—especially in low- and middle-income settings—governments are increasingly seeking to adopt analytics-driven approaches to public sector human resource (HR) management and civil service reform. Yet most government offices and our World Bank regional counterparts often lack standardized tools for transforming raw HRMIS data into formats suitable for rigorous analysis. As a result, each new diagnostic requires bespoke data cleaning, ad hoc coding decisions, and country-specific scripts that are costly to maintain and difficult to replicate.

This fragmentation makes HRM analytics expensive, slow, and inconsistent. It also limits the ability of governments and development partners—including the World Bank—to compare results across time, sectors, or countries, or to build reusable analytics pipelines.

This article describes the standard approach for harmonizing human resource management information system (HRMIS) data, organized into three main modules: Establishment, Personnel, and Contract. The contents of each module are described in `vignette("standard_dictionary")`. In this article, we provide a set of helper functions to support the harmonization process across the modules. When users prepare their HRMIS data according to this standard, teams can immediately apply automated tools for data quality control as well as dynamic and reproducible reports on workforce structure dynamics, pay and compensation analytics, establishment control monitoring, staffing distribution analysis. 

This vignette, harmonization.Rmd, presents the practical workflow for transforming raw HRMIS extracts into these standardized modules. It provides governments with a clear, repeatable procedure for preparing their data so that advanced analytics can be deployed rapidly and consistently—reducing costs, increasing comparability, and strengthening evidence-based HR decision making.

The harmonization workflow demonstrated here shows how to:

* map raw administrative HR variables into a standardized public sector schema,

* recode local classifications (education, occupation, contract type) into cross-country taxonomies,

* generate unique and persistent identifiers for personnels, establishments, and contracts,

* attach missing metadata such as country codes, administrative hierarchies, or reference dates,

* structure the cleaned data into the three standard HRMIS modules, and

* validate the outputs against the HRMIS Standard Data Dictionary.

By following this procedure, governments and World Bank teams can unlock a common analytics ecosystem in which standardized public sector HR data feeds directly into automated dashboards, diagnostic tools, and monitoring systems—enabling faster, cheaper, and more consistent evidence-based HR reform.

## The Raw Data


For this vignette, we apply our tools to data provided by Brazil HRMIS system provided at the contract level. We randomly sample 5 percent of the original data. See the data below:

```{r dataview}
bra_hrmis <- govhr::bra_hrmis |> 
  sample_frac(0.05)

reactable(
  head(bra_hrmis, 1e3)
)

```

We illustrate the harmonization workflow using this dataset. `bra_hrmis` contains `r nrow(bra_hrmis)` rows and `r ncol(bra_hrmis)` columns, covering covering demographic, employment, establishmental, and payroll information for public sector personnels.

A quick glimpse of the dataset: 

```{r glimpsedata}

glimpse(bra_hrmis)

```

This will give you a brief, surface level overview of the dataset. You should see the dimensions of the data (i.e. number of rows and columns) and variable types and perhaps begin to get a sense for completeness (or the lack thereof) for certain variables. 


# The Harmonization Process

## Introduction  
The goal of the harmonization process is to establish a consistent and standardized structure for cleaned HR datasets, ensuring that analytics supporting government human resources management from a common analytical foundation. In many client countries, HRM analytics are conducted in an ad-hoc, highly customized manner that varies by consultant, ministry, year, and even dataset. This leads to high analytical costs, limited comparability over time or across institutions, and significant barriers to scaling evidence-based human resource reform.

The harmonization framework addresses these challenges directly. By defining a clear data dictionary and a set of standardized transformations, governments can prepare their HRMIS data in a format that is analytically ready and fully compatible with the tools provided in the `govhr` package. This ensures:

- **Reproducible and transparent workflows**  
- **Cross-country and cross-institution comparability**  
- **Automatic generation of quality checks and dashboards**  
- **Reduced data preparation cost for each new engagement**  

The remainder of this section documents the harmonization steps for each of the three core modules derived from the data dictionary:

1. **Contract Module**  
2. **Personnel Module**
3. **Establishment Module**
Each module represents a specific level of analysis defined by the dictionary and is to be constructed directly from the raw HRMIS data.

---

## Harmonizing the Contract Module

The Contract Module is the foundational component of the harmonized payroll dataset. The purpose of this module is to extract and structure all information that is **unique at a contract, reference date level**, as defined in the data dictionary. A “contract record” refers to a personnel–contract relationship at a specific reference period (usually month–year), capturing the contractual characteristics governing the personnel's employment at that time.

In practice, the raw HRMIS data typically mixes variables at different conceptual levels—personnel attributes, position attributes, contract terms, payroll events, and sometimes even one-off administrative transactions. 

Working with payroll or HRMIS data in many developing-country contexts often involves very large datasets—sometimes millions of records spanning multiple years. Such data volumes can quickly overwhelm standard analytical workflows if not handled using efficient tools.

In R, there are several paradigms for manipulating two-dimensional data. For the purposes of this tutorial, we rely on the data.table package for the tasks that are more computationally intensive. At the time of writing, data.table remains the gold standard for high-performance data processing in R, offering exceptional speed, efficient memory use, and an expressive syntax tailored for large datasets.

Although the dplyr package from the tidyverse ecosystem is widely used and favored for its readability and intuitive grammar, operations on large payroll datasets can be substantially slower when using dplyr alone. To ensure reproducibility, performance, and scalability—especially for teams with limited computational resources—we adopt data.table as the primary engine for harmonization. However, we will also use the dplyr functions where performance is not improved to ensure code readability. 

The `bra_hrmis` data is written in Portuguese which might require translation to English to ease understanding. One option is to use an LLM like ChatGPT to get a quick as there are only `r length(colnames(bra_hrmis))` columns. Alternatively the `polyglotr` R package provides a suite functions that support translation. This can be applied as follows:

```{r colname-dictionary}

raw_dictionary <- 
  tibble(raw_colnames_pt = colnames(govhr::bra_hrmis),
         raw_colnames_eng = polyglotr::google_translate(colnames(bra_hrmis), "pt", "en"))

kable(raw_dictionary)

```


The first step when handling any raw payroll dataset is to find the individual to identifies each countract for each time period unique. We begin by looking through the `bra_hrmis` object for this. From the above table and the previous glimpse(), we can see that the `CPF` and `MATRICULA` variables are the most likely candidates to identify contracts uniquely. We do a quick check for this as follows:

```{r findunique}

### first let us convert the data to a data.table object to speed up our computations

bra_hrmis <- as.data.table(bra_hrmis)

# Unique CPF per year
cpf_summary <- bra_hrmis[, .(unique_cpf = uniqueN(CPF),
                             nobs       = .N),
                         by = ANO_PAGAMENTO]

# Unique Matricula per year
mtr_summary <- bra_hrmis[, .(unique_mtr = uniqueN(MATRICULA),
                             nobs       = .N), 
                         by = ANO_PAGAMENTO]

kable(cpf_summary)

kable(mtr_summary)

```

This clearly shows that `MATRICULA` is the contract ID while `CPF` is possibly the identifier for the personnel, the latter will come in handy during the harmonization of the personnel module. 

Now, we can begin creating the set of variables as defined by the `standard_dictionary` vignette. It is often useful to begin by creating the derived variables. There are four sets of derived variables within the dictionary: 

 * (1) the Industrial Standard Classification of Occupation (ISCO) variables, i.e. `occupation_isconame`, `occupation_iscocode` are derived from the original occupation variables `occupation_native`, `occupation_english`. We apply the `polyglotr::google_translate()` function to convert Portuguese named occupations to English as well as use the `LabourR::classify_occupation()` function to classify these occupations to the level-4 isco names. 
 
```{r harmonize-occuapation, warning=FALSE}


#-----------------------------
# 2. Build occupation table (active + inactive)
#-----------------------------

occup_df <- bra_hrmis[, c("CARREIRA", "CARGO")]

# distinct CARREIRA/CARGO/status
occup_df <- unique(occup_df)

# Add translated vars
occup_df[, occupation_native := tolower(CARREIRA)]

occup_df[, occupation_english := tolower(polyglotr::google_translate(text = occupation_native,
                                                          source_language = "pt",
                                                          target_language = "en"))]
```

```{r}
class_occup_df <- copy(occup_df)[,
  .(id = .I, text = occupation_english)
]

class_occup_df <- classify_occupation(
  class_occup_df,
  isco_level = 4,
  lang = "en",
  num_leaves = 1
)
```
```{r, echo = FALSE}
reactable(
  head(
    class_occup_df, 
    100
  )
) 
```
 
```{r merge-harmonized-occupation}


#-----------------------------
# 4. Merge classification back into occup_df
#-----------------------------
occup_df[, id := .I]

# merge iscoGroup
occup_df <- merge(
  occup_df,
  class_occup_df[, .(id = as.integer(id), iscoGroup)],
  by = "id",
  all.x = TRUE
)

setnames(occup_df, "iscoGroup", "occupation_iscocode")

# merge ISCO descriptions
occup_df <- merge(occup_df,
                  isco[, c("unit", "description")] |> as.data.table(),
                  by.x = "occupation_iscocode",
                  by.y = "unit",
                  all.x = TRUE)

setnames(occup_df, "description", "occupation_isconame")

#-----------------------------
# 5. Bring classified occupations back to bra_hrmis (active only)
#-----------------------------

## we perform a left join with data.table syntax for speed as the merging into 
## the hrmis dataset could be computationally intensive

bra_hrmis <- occup_df[bra_hrmis, on = c("CARREIRA", "CARGO")]

```
  * (2) the date variables (`ref_date`, `start_date`, `end_date`) are often in the serial 5-digit format and need to be standardized as they will be used in preparing the compensation variables which are time-variant. Below, we create the date variables:
  
```{r harmonize-dates, warning = FALSE, message = FALSE}

### lets include the dates
bra_hrmis[, ref_date := as.Date(paste(ANO_PAGAMENTO, MES_REFERENCIA, "01", sep = "-"))]

setnames(bra_hrmis, c("DATA_ADMISSAO", "DATA_APOSENTADORIA"), c("start_date", "end_date"), skip_absent = TRUE)
bra_hrmis[, start_date := as.Date(as.integer(start_date), origin = "1899-12-30")]
bra_hrmis[, end_date := as.Date(as.integer(end_date), origin = "1899-12-30")]


### lets include the country code and admin identifier

bra_hrmis[, country_code := "BRA"]
bra_hrmis[, country_name := "Brazil"]
bra_hrmis[, adm1_name := "Alagoas"]
bra_hrmis[, adm1_code := "AL"]

```

   
   
   
  * (3) the compensation variables i.e. `gross_salary_ppp`, `base_salary_ppp`, `net_salary_ppp` which are all derived from the local currency equivalents `gross_salary_lcu`, `base_salary_lcu`, `net_salary_lcu`. We provide a function called `convert_constant_ppp()` as well as a `macro_indicators` dataset with this package to perform these transformations between the nominal variables (i.e. _lcu) and their real equivalents (_ppp). 
 
 
```{r harmonize-compensation, warning = FALSE, message = FALSE}

### lets convert the nominal compensation variables to real (_ppp) values
# Step 1: Create the _lcu salary variables (in-place, no copy)

### lets rename the raw variables here to their actual names
setnames(bra_hrmis, 
         old = c("SALARIO_BASE", "SALARIO_BRUTO", "SALARIO_LIQUIDO", "ABONO_PERMANENCIA"),
         new = c("base_salary_lcu", "gross_salary_lcu", "net_salary_lcu", "allowance_lcu"))


# Step 2: Identify all *_lcu columns to convert
cols_to_convert <- grep("_lcu$", names(bra_hrmis), value = TRUE)

# Step 3: Apply PPP conversion using your convert_constant_ppp function
pfw_df <- 
  macro_indicators |>
  dplyr::filter(country_code == "BRA", 
                year %in% as.integer(unique(bra_hrmis$ANO_PAGAMENTO))) |>
  dplyr::select(all_of(c("country_code", "year", "cpi", "ppp")))


bra_hrmis[, (cols_to_convert) := lapply(.SD, as.numeric), .SDcols = cols_to_convert]


### apply the convert_constant_ppp function to produce those estimates
bra_hrmis <- convert_constant_ppp(data = bra_hrmis[, year := lubridate::year(ref_date)],
                                  cols = cols_to_convert,
                                  macro_indicators = macro_indicators)

### while we are at this lets include the working hour variable as well
bra_hrmis[, whours := as.numeric(JORNADA)]

```


 * (4) the contract type variable `TIPO_CONTRATO` needs to be reclassified according to the dictionary, `contract_type`. We do so as follows:
 
```{r harmonize-contracttype}

### create a little dictionary mapping all the raw classes into a contract type in a data.table
contract_dict <- 
  data.table(TIPO_CONTRATO = unique(bra_hrmis$TIPO_CONTRATO),
             contract_type = c("short-term", "permanent", "permanent", "open-term",
                               "short-term", "inactive", "retired"))

kable(contract_dict)
```
 
 
Let's do a data.table join of the `contract_dict` into the `bra_hrmis`
```{r merge-contract-type}

bra_hrmis <- contract_dict[bra_hrmis, on = "TIPO_CONTRATO"]


```


Now that the derived variables in this contract module have been created, we are ready to create the rest of the variables all at once to finalize the module. 

```{r finalize-contract-harmonization}

setnames(bra_hrmis,
         old = c("MATRICULA", "CPF", "ORGAO", "CLASSE", "NIVEL"),
         new = c("contract_id", "personnel_id", "est_id", "paygrade", "seniority"))


### now let us select the final set of dictionary variables to complete the module
### lets use dplyr::select() function since this will not add any computational time
bra_hrmis_contract <- 
  bra_hrmis |>
  dplyr::select(
    contract_id, personnel_id, est_id,
    ends_with("_date"),
    contains("_salary_"),
    contract_type, 
    starts_with("occupation_"),
    country_code, country_name,
    adm1_name, adm1_code,
    whours, paygrade, seniority
  )


```

Here is what our final data looks like:
```{r output-final-contract-harmonization, echo = FALSE}

reactable(
  head(bra_hrmis_contract, 100)
)

```


## Harmonizing the Personnel Module

The Personnel Module standardizes individual-level information about employees with the public sector. This section demonstrates the harmonization pipeline using the same contract-level Brazil (Alagoas) HRMIS dataset (bra_hrmis) as an example. We will produce a clean, harmonized dataset named bra_hrmis_personnel which conforms to the Personnel Module dictionary in `standard_dictionary.Rmd`.

Let's start with a little exploration. We need to understand how many unique personnel-establishment-refdate combinations are within the data:

```{r creating-unique-personnel-identifiers}

## how many observations should we expect to have
personnel_count <- 
bra_hrmis |>
  dplyr::select(personnel_id, est_id, ref_date) |>
  uniqueN()

```

This tells us that we have `r personnel_count` unique personnel_id-est_id-refdate combinations in the entire dataset. The goal is to ensure that once all the other variables of the personnel module are added. The size of the personnel module remains the same. As we did, in the contract module, we now add all the derived variables:

```{r add-derivative-variables}

## lets relabel some more variables


# Create the education harmonization dictionary for educat7
education_dictionary <- data.table(
  ESCOLARIDADE = c(
    "ANALFABETO",
    "1 A 4 SERIE DO PRIM. GRAU INCOMPLETO",
    "5 A 8 SERIE DO PRIM. GRAU INCOMPLETO",
    "1 A 4 SERIE DO PRIM. GRAU COMPLETO",
    "5 A 8 SERIE DO PRIM. GRAU COMPLETO",
    "SEGUNDO GRAU INCOMPLETO",
    "SEGUNDO GRAU COMPLETO",
    "ESPECIALIZAÇÃO COMPLETO",
    "ESPECIALIZAÇÃO INCOMPLETO",
    "ESPECIALIZA«√O COMPLETO",
    "ESPECIALIZA«√O INCOMPLETO",
    "CURSO SUPERIOR COMPLETO",
    "CURSO SUPERIOR INCOMPLETO",
    "MESTRADO INCOMPLETO",
    NA_character_
  ),
  educat7 = c(
    "No education",
    "Primary incomplete",
    "Primary incomplete",
    "Primary complete",
    "Primary complete",
    "Secondary incomplete",
    "Secondary complete",
    "Higher than secondary but not university",
    "Higher than secondary but not university",
    "Higher than secondary but not university",
    "Higher than secondary but not university",
    "University incomplete or complete",
    "University incomplete or complete",
    "University incomplete or complete",
    NA_character_
  )
)

# Merge the dictionary into bra_hrmis
bra_hrmis <- education_dictionary[bra_hrmis, on = "ESCOLARIDADE"]


bra_hrmis <- 
bra_hrmis |>
  setnames(old = c("DATA_NASCIMENTO", "GENERO"),
           new = c("birth_date", "gender"))

### lets prepare the birth date variables
bra_hrmis[, birth_date := as.Date(as.integer(birth_date), origin = "1899-12-30")]

```

The remainder of the variables (`tribe`, `race``) appear to be missing from the `bra_hrmis` raw data. Therefore, we create them as missing variables (`NA`). 

```{r}

bra_hrmis[, c("tribe", "race") := .(NA, NA)] ## lets quickly create the variables that are missing from the raw data

```


Finally, we can create the personnel module by taking the set of unique values across the set of personnel modules we have now created within `bra_hrmis`. See the implementation below:

```{r creating-unique-personnel-identifiers2}


bra_hrmis_personnel <- 
  bra_hrmis |>
  dplyr::select(personnel_id, est_id, ref_date, birth_date, gender, educat7,
                country_name, country_code, adm1_name, adm1_code) |>
  unique()


```

Let's take a look at the `bra_hrmis_personnel`

```{r, echo = FALSE}

reactable(
  head(bra_hrmis_personnel, 100)
)

```


## Harmonizing the Establishment Module

The Establishment Module extracts, standardizes, and structures information on public-sector establishments from the HRMIS records, `bra_hrmis`. The steps below convert the raw establishment identifiers into a canonical, well-structured establishment register according the harmonization dictionary. See below: 

```{r establishment-module}

### get the set of variables according to the dictionary
bra_hrmis_est <- 
  bra_hrmis |>
  mutate(est_name_native = est_id,
         est_type = NA,
         est_parent = NA,
         est_child = NA) |>
  dplyr::select(
    est_id, est_name_native, ref_date, est_type, est_parent, est_child,
    country_code, country_name, adm1_name, adm1_code
  ) |>
  unique() |>
  head(10) |> 
  mutate(est_name_en = polyglotr::google_translate(est_name_native, target_language = "en"))

```

```{r, echo = FALSE}
reactable(
  head(bra_hrmis_est, 10)
)
```
