---
title: "Data harmonization"
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(govhr)
library(dplyr)
library(data.table)
```

# Why harmonize?

Governments collect HR data in their own conventions: column names, classifications, and identifiers rarely match a common standard. Harmonization re-expresses raw HR data extractions using the `govhr` [Standard Data Dictionary](https://wb-pida-data-science-shop.github.io/govhr/articles/01-standard_dictionary.html), organized into three modules: **Establishment**, **Personnel**, and **Contract**. Once harmonized, data becomes comparable across countries and over time, and can feed directly into `govhr`'s analytics functions.

This article walks through harmonizing a synthetic dataset, `bra_hrmis`, built from a contract-level extract from Brazil's HRMIS system in the Alagoas state, using the helper functions in `govhr`. The example focuses primarily on generating one module, Personnel, but a similar logic can be applied to the Establishment and Contract modules.

```{r}
bra_hrmis <- as.data.table(govhr::bra_hrmis)
dim(bra_hrmis)
```

# Step 1: check column consistency

Before harmonizing, it's useful to check whether raw data sources (e.g. multiple years or offices) share a consistent column naming (i.e., schema). `find_inconsistent_colnames()` flags columns that aren't shared across all sources.

```{r}
# simulate two extracts with slightly different schemas
extract_2020 <- bra_hrmis[ANO_PAGAMENTO == min(ANO_PAGAMENTO)]
extract_2021 <- bra_hrmis[ANO_PAGAMENTO == max(ANO_PAGAMENTO)]

find_inconsistent_colnames(list(extract_2020, extract_2021))
```

# Step 2: generate module identifiers

Harmonization requires knowing which column identifies a contract and which identifies a person. These are our primary keys, that is, unique identifiers for contracts and personnel, respectively. `find_duplicate_ids()` is a quick way to test whether a candidate column is a valid unique key.

```{r}
bra_hrmis |> find_duplicate_ids(MATRICULA)
```

`MATRICULA` uniquely identifies a contract for a given reference date, so it becomes the **contract identifier**, or `contract_id` in our standard data dictionary. `CPF` will serve as the **personnel identifier**, the `personnel_id`.

# Step 3: map raw columns to standardized names

Raw data columns rarely match the naming conventions of the `govhr` standard dictionary. `harmonize_columns()` renames the data with a key-value dictionary, keeping only the desired columns.

```{r}
column_dictionary <- c(
  contract_id  = "MATRICULA",
  personnel_id = "CPF",
  est_id       = "ORGAO",
  paygrade     = "CLASSE",
  seniority    = "NIVEL",
  gender       = "GENERO"
)

bra_hrmis <- harmonize_columns(bra_hrmis, column_dictionary) |>
  cbind(bra_hrmis[, !names(column_dictionary), with = FALSE][
    ,
    setdiff(names(bra_hrmis), names(column_dictionary)),
    with = FALSE
  ])
```

# Building reference dates

Dates are often stored as serial numbers or split across year/month columns. Once a `ref_date` column is built, `guess_date_frequency()` confirms the reporting interval — useful for choosing the right window when computing growth or turnover indicators later.

```{r}
bra_hrmis[, ref_date := as.Date(paste(ANO_PAGAMENTO, MES_REFERENCIA, "01", sep = "-"))]

bra_hrmis |>
  as.data.frame() |>
  guess_date_frequency()
```

# Deduping values

Attributes like education and gender should be constant for a given personnel-period, but raw data sometimes disagrees across rows for the same `personnel_id`/`ref_date` (e.g. due to concurrent contracts). `dedup_values()` resolves this using a chosen strategy — here, the modal value.

```{r}
bra_hrmis <- dedup_values(
  bra_hrmis,
  id_col = personnel_id,
  date_col = ref_date,
  value_col = gender,
  method = "mode"
) |>
  select(personnel_id, ref_date, gender) |>
  right_join(
    bra_hrmis |> select(-gender),
    by = c("personnel_id", "ref_date")
  ) |>
  as.data.table()
```

# Deflating wages

Nominal wages in local currency aren't comparable across time or countries. `convert_constant_ppp()` uses the bundled `macro_indicators` dataset (with CPI and PPP conversion factors) to convert local currency units into international dollars.

```{r}
setnames(
  bra_hrmis,
  old = c("SALARIO_BASE", "SALARIO_BRUTO", "SALARIO_LIQUIDO"),
  new = c("base_salary_lcu", "gross_salary_lcu", "net_salary_lcu")
)

bra_hrmis[, country_code := "BRA"]

salary_cols <- grep("_lcu$", names(bra_hrmis), value = TRUE)
bra_hrmis[, (salary_cols) := lapply(.SD, as.numeric), .SDcols = salary_cols]

bra_hrmis <- convert_constant_ppp(bra_hrmis, cols = salary_cols)

bra_hrmis |>
  select(personnel_id, ref_date, all_of(salary_cols), ends_with("_ppp")) |>
  head()
```

`deflate_to_real()` offers a lighter-weight alternative when you only need compensation expressed in constant local currency units (rather than constant international dollars):

```{r}
bra_hrmis[, gross_salary_real := deflate_to_real(
  gross_salary_lcu,
  ref_date,
  country_code,
  base_year = 2021
)]
```

# Completing the module

Not every data extraction contains every dictionary column (e.g. `tribe`, `race` may be unavailable). `complete_columns()` adds any missing dictionary columns as `NA`, so the resulting table always matches the expected schema — no more, no less.

```{r}
personnel_cols <- c(
  "personnel_id", "est_id", "ref_date", "gender",
  "educat7", "tribe", "race", "country_code"
)

bra_hrmis_personnel <- bra_hrmis |>
  select(any_of(personnel_cols)) |>
  distinct() |>
  complete_columns(personnel_cols)

bra_hrmis_personnel |> head()
```

# Validating the harmonized data

Once harmonized, `validate_data()` checks the harmonized dataset against a set of validation rules. It then reports both a pass-rate summary and the specific records that violate each rule. This helps you identify negative salaries, missing keys, or out-of-range dates before analytics are run.

```{r}
result <- validate_data(
  data        = govhr::bra_hrmis_contract,
  input_rules = govhr::contract_rules
)

result$report
```

You can inspect the specific records that violate a given rule, for example, the `salary_non_negative` rule:

```{r}
result$violations[["salary_non_negative"]]
```

Informed by this validation, you might want to revisit your harmonization process and address lingering data quality issues.
