---
title: "HR Analytics"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  dpi = 300
)

library(dplyr)
library(ggplot2)
library(reactable)
library(ggthemes)
library(data.table)
library(stringr)
library(govhr)

bra_hrmis_contract <- govhr::bra_hrmis_contract |> 
  convert_constant_ppp(
    cols = "net_salary_lcu",
    macro_indicators = macro_indicators
  ) |> 
  mutate(
    org_id = str_extract(org_id, "^[^-]+")
  ) |> 
  as.data.table()

bra_hrmis_contract[, year := lubridate::year(ref_date)]

theme_set(
  theme_few()
)
```

# Overview

This article shows examples for how to analyze harmonized human resource management information system (HRMIS) data, organized into three main modules: Organization, Worker, and Contract. The contents of each module are described in `vignette("standard_dictionary")`

# Building analytics on harmonized data

Because we have already cleaned and harmonized the HRMIS data shared by the Government of Brazil, we can use standard functions to produce analytical outputs. This is fast and reliable because the function knows what to data formats to expect and produces summary statistics in a consistent way. 

```{r, message=FALSE, warning=FALSE}
library(govhr)
```

Let's start with the contract module. Suppose we are interested in computing the overall wage bill and its evolution over time. We can do it quickly with this convenient function:

```{r}
wagebill_annual <-
  bra_hrmis_contract |> 
    compute_fastsummary(
      cols = c("net_salary_ppp"), 
      fns = c("sum"), 
      groups = c("year"), 
      output = "long"
    )

reactable(
  wagebill_annual
)
```

There is no salary data for the year of 2018, therefore we can't compute the wage bill for that year.

We can quickly visualize the annual wage bill.

```{r}
#| fig.alt: >
#|   Plot of wage bill at the annual level.

wagebill_annual |> 
  filter(
    year != 2018
  ) |> 
  ggplot_point_line(
    year,
    value
  ) +
  scale_y_continuous(
    labels = scales::label_number()
  ) +
  labs(
    x = "Year",
    y = "Wage Bill \n(Constant International Dollars)",
    title = "Wage Bill: Annual"
  )
```

Now imagine you want to unpack the wage bill. Suppose you are interested in a decomposition by establishment. You can quickly do that with our function, like this:

```{r}
wagebill_annual_org <-
  bra_hrmis_contract |> 
    compute_fastsummary(
      cols = c("net_salary_ppp"), 
      fns = c("sum"), 
      groups = c("org_id", "year"), 
      output = "long"
    )

reactable(
  wagebill_annual_org
)
```

We just have to add `"org_id"` to the groups argument. We can quickly plot this new object:

```{r}
#| fig.alt: >
#|   Plot of wage bill at the annual level and organizational level.

wagebill_annual_org |> 
  filter(
    year != 2018 &
      org_id %in% (unique(wagebill_annual_org$org_id) |> sample(4))
  ) |>
  ggplot_point_line(
    year,
    value,
    group = org_id
  ) +
  scale_y_continuous(
    labels = scales::label_number()
  ) +
  facet_wrap(
    vars(org_id)
  ) +
  labs(
    x = "Year",
    y = "Wage Bill \n(Constant International Dollars)",
    title = "Wage Bill: Annual, by Organization"
  )
```

