
---
title: "Quality Control Report"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(govhr)
```

## Overview

This article demonstrates the quality control reporting capabilities of the `govhr` package using the Brazilian HRMIS synthetic dataset.

## Generate Quality Control Report

The `generate_qc_report()` function creates a comprehensive HTML report that includes:

- **Data Basics**: Module dimensions, temporal coverage, variable structure, primary key integrity, and cross-module orphan checks
- **Missingness Report**: Visualizations of missing data patterns by occupation, ISCO code, time period, and establishment
- **Volatility Analysis**: Heatmaps showing temporal stability of salary, contract counts, and work hours
- **Q-Check Summary**: Pointblank validation checks for structure, keys, orphans, salaries, dates, missingness, and volatility

```{r generate-report, eval = FALSE}
# Generate the quality control report
generate_qc_report(
  contract_dt = bra_hrmis_contract,
  personnel_dt = bra_hrmis_personnel,
  est_dt = bra_hrmis_est,
  output = "quality_control_report.html"
)
```

```{r actually-generate, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE}
# Actually generate the report for the pkgdown site
tryCatch({
  output_path <- file.path("articles", "quality_control_report.html")
  
  generate_qc_report(
    contract_dt = bra_hrmis_contract,
    personnel_dt = bra_hrmis_personnel,
    est_dt = bra_hrmis_est,
    output = output_path
  )
  
  report_generated <- TRUE
}, error = function(e) {
  message("Report generation skipped: ", e$message)
  report_generated <- FALSE
})
```

```{r show-link, echo = FALSE, results = 'asis', eval = exists("report_generated") && report_generated}
cat("\n---\n\n**[View the full Quality Control Report →](quality_control_report.html)**\n")
```