---
title: "Standard HR Data Quality Control Report (Q-Check)"
author: "Ifeanyi Edochie"
date: "`r Sys.Date()`"
params:
  contract_dt: NULL
  personnel_dt: NULL
  est_dt: NULL
  qc_obj: NULL
  run_qc: TRUE
    
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(gt)
library(data.table)
library(naniar)
library(viridis)


```

# Standard HR Quality Control Analytics Report 

## Overview

This report presents a comprehensive quality control assessment of the harmonized payroll data comprising the **Contract**, **Personnel**, and **Establishment** modules according to the standard harmonization dictionary (`govhr::harmonization_dict()`). All diagnostic tables are generated automatically using the `compute_qualitycontrol()` function, which evaluates the aforementioned harmonized modules .

The quality control report is subdivided as follows:

- **Data Basics**  
  High-level statistics summarizing the size, structure, and key characteristics of harmonized data.

- **Variable Structure & Dictionary Conformity**  
  Comparison of each module’s variable names against the harmonization dictionary to identify missing, extra, or mismatched fields.

- **Primary Key Integrity**  
  Validation that the Contract module’s identifying keys uniquely and consistently describe individual records.

- **Cross-Module Orphan Checks**  
  Verification that all referenced `personnel_id` and `est_id` values in the Contract module exist in the corresponding Personnel and Establishment modules and vice versa.

- **Salary Logic Checks**  
  Screening for numeric inconsistencies, negative values, and violations of expected salary relationships (e.g., base > gross, net > gross).

- **Date Logic Validation**  
  Detection of invalid employment record sequences, such as end dates occurring before start dates.

- **Missingness Diagnostics**  
  Measurement and visualization of missing data overall and across key analytical dimensions (occupation, ISCO mapping, reference period, and establishment).

- **Volatility and Stability Metrics**  
  Time-series diagnostics evaluating changes in salary aggregates, wagebill, staff counts, and contract counts across reference periods.

A final section consolidates all checks into a unified **pointblank** summary report, providing a high-level view of which components pass, fail, or warrant further attention.


## Diagnostic Sections {.tabset .tabset-pill}

### Data Basics 

This section provides a high-level profile of the harmonized payroll data and summarizes the structure of the three core modules—**Contract**, **Personnel**, and **Establishment**—before detailed diagnostics are introduced.

#### Module Dimensions

The harmonized dataset includes the following components:

```{r, echo = FALSE}


dim_dt <- data.table(
  Module = c("Contract", "Personnel", "Establishment"),
  Observations = c(
    params$qc_obj$n_obs,
    if (!is.null(params$personnel_dt)) nrow(params$personnel_dt) else NA,
    if (!is.null(params$est_dt)) nrow(params$est_dt) else NA
  ),
  Variables = c(
    params$qc_obj$n_vars,
    if (!is.null(params$personnel_dt)) ncol(params$personnel_dt) else NA,
    if (!is.null(params$est_dt)) ncol(params$est_dt) else NA
  )
)

dim_dt |>
  gt() |>
  tab_header(
    title = md("### Module Dimensions Summary")
  ) |>
  fmt_number(
    columns = c(Observations, Variables),
    use_seps = TRUE,
    decimals = 0
  ) |>
  cols_align(
    align = "center",
    columns = c(Observations, Variables)
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#f7f7f7")
    ),
    locations = cells_body()
  ) |>
  opt_table_outline() |>
  opt_row_striping()


```
  

These modules are expected to integrate cleanly through common identifiers (`personnel_id`, `est_id`, and `ref_date`), and later sections evaluate the extent to which these linkages hold.

## Missingness Report

This section examines patterns of missing data across the Contract, Personnel, and Establishment modules.  
Understanding missingness is essential for data quality control because it highlights variables that may require cleaning, imputation, or further validation by the data provider.

We evaluate:

1. **Variable-level missingness** (percent missing in each variable)  
2. **Module-level summaries**  
3. **Visualizations**  
   - Missingness barplots  
   - Missingness heatmaps  


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=6}

overall_dt <- params$qc_obj$missingness$overall

ggplot(overall_dt, aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Overall Missingness by Variable",
    x = "Variable",
    y = "Percent Missing"
  ) +
  theme_minimal(base_size = 13)


```


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=7}

miss_occ <- params$qc_obj$missingness$by_occupation

# pick top occupations by missingness severity
top_occ <- miss_occ[
  , .(mean_missing = mean(pct_missing)), by = occupation_native
][order(-mean_missing)][1:30, occupation_native]

plot_dt <- miss_occ[occupation_native %in% top_occ]

ggplot(plot_dt, aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ occupation_native, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Missingness by Variable — Top 30 Occupations",
    x = "Variable",
    y = "Percent Missing"
  ) +
  theme_minimal(base_size = 11)

```



```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=7}

miss_isco <- params$qc_obj$missingness$by_isco

top_isco <- miss_isco[
  , .(mean_missing = mean(pct_missing)), by = occupation_isconame
][order(-mean_missing)][1:30, occupation_isconame]

plot_dt <- miss_isco[occupation_isconame %in% top_isco]

ggplot(plot_dt, aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ occupation_isconame, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Missingness by Variable — Top 30 ISCO Groups",
    x = "Variable",
    y = "Percent Missing"
  ) +
  theme_minimal(base_size = 11)

```


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=6}

miss_ref <- params$qc_obj$missingness$by_ref
  
ggplot(miss_ref, aes(x = factor(lubridate::year(ref_date)), y = variable, fill = pct_missing)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_viridis(option = "C", labels = scales::percent_format()) +
  labs(
    title = "Missingness Heatmap by Reference Date",
    x = "Reference Date",
    y = "Variable",
    fill = "% Missing"
  ) +
  theme_minimal(base_size = 12)

```



```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=7}

miss_est <- params$qc_obj$missingness$by_est

top_est <- miss_est[
  , .(mean_missing = mean(pct_missing)), by = est_id
][order(-mean_missing)][1:30, est_id]

plot_dt <- miss_est[est_id %in% top_est]

ggplot(plot_dt, aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ est_id, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Missingness by Variable — Top 30 Establishments",
    x = "Variable",
    y = "Percent Missing"
  ) +
  theme_minimal(base_size = 10)

```
