---
title: "Standard HR Report"
output:
  html_document:
    toc: true
params:
  workforce: NULL
  wagebill: NULL
---

```{r setup, include = FALSE}
# the indicators are computed by generate_standard_report(); this template
# only draws them
library(dplyr)
library(ggplot2)
library(govhr)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5
)

theme_set(theme_minimal(base_size = 13))

workforce <- params$workforce
wagebill <- params$wagebill

# the five establishments with the most staff on the latest date, shown in
# every chart by establishment so the charts stay readable
top_establishments <- workforce$headcount_by_est |>
  filter(ref_date == max(ref_date)) |>
  slice_max(headcount, n = 5, with_ties = FALSE) |>
  pull(est_id)
```

# Workforce

## Headcount

Number of people working in government on each date.

```{r headcount}
workforce$headcount |>
  plot_trend(group_col = "ref_date", y_col = "headcount", y_label = "Headcount")
```

## Headcount by establishment

The five establishments with the most staff on the latest date.

```{r headcount-by-est}
workforce$headcount_by_est |>
  filter(est_id %in% top_establishments) |>
  plot_trend(
    group_col = "est_id",
    y_col = "headcount",
    y_label = "Headcount"
  ) +
  labs(color = "Establishment") +
  theme(legend.position = "bottom")
```

## Hires and separations

Hires are people active on a date but not on the previous one; separations
are people active on a date but not on the next one, including retirements.
Both are shown as a share of headcount.

```{r hires}
plot_movement(
  workforce$movement,
  movement_type = "hire",
  measurement_type = "rate"
)
```

```{r separations}
plot_movement(
  workforce$movement,
  movement_type = "separation",
  measurement_type = "rate"
)
```

## Transitions

Total number of transitions between establishments by personnel.

```{r transitions}
if (nrow(workforce$transitions) == 0) {
  cat("No moves between establishments were recorded.")
} else {
  # Word cannot show interactive charts, so draw a static network there
  plot_transition_network(
    workforce$transitions,
    interactive = knitr::is_html_output()
  )
}
```

# Wagebill

## Total wagebill

Total pay across all contracts on each date.

```{r wagebill}
wagebill$wagebill |>
  plot_trend(group_col = "ref_date", y_col = "wagebill", y_label = "Wagebill")
```

## Wagebill growth

Change in the wagebill since the previous date.

```{r wagebill-growth}
wagebill$wagebill |>
  filter(!is.na(wagebill_growth)) |>
  plot_trend(
    group_col = "ref_date",
    y_col = "wagebill_growth",
    y_label = "Growth"
  ) +
  scale_y_continuous(labels = scales::label_percent()) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50")
```

## Wagebill by establishment

The wagebill of the five establishments with the most staff on the latest
date.

```{r wagebill-by-est}
wagebill$wagebill_by_est |>
  filter(est_id %in% top_establishments) |>
  plot_trend(group_col = "est_id", y_col = "wagebill", y_label = "Wagebill") +
  labs(color = "Establishment") +
  theme(legend.position = "bottom")
```

## Average wage

Average pay per contract on each date.

```{r wage}
wagebill$wage |>
  plot_trend(group_col = "ref_date", y_col = "wage", y_label = "Average wage")
```

## Pay composition

Share of the wagebill paid through each pay component.

```{r composition}
wagebill$composition |>
  ggplot(aes(ref_date, share_wagebill, fill = component)) +
  geom_area() +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(x = "Time", y = "Share of wagebill", fill = "Component") +
  theme(legend.position = "bottom")
```

## Pay distribution

Share of contracts in each pay band, pooling all dates.

```{r distribution}
plot_histogram(wagebill$distribution)
```
