---
title: "Standard HR Report"
output:
  html_document:
    toc: true
params:
  workforce: NULL
  wagebill: NULL
---

```{r setup, include = FALSE}
# the indicators are computed by generate_standard_report(); this template
# only draws them
library(dplyr)
library(ggplot2)
library(govhr)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5
)

theme_set(theme_minimal(base_size = 13))

workforce <- params$workforce
wagebill <- params$wagebill

# keep the `n` establishments with the largest `value_col` on the latest date,
# so charts by establishment stay readable
top_establishments <- function(data, value_col, n = 5) {
  latest <- data |>
    filter(ref_date == max(ref_date), !is.na(.data[[value_col]])) |>
    slice_max(.data[[value_col]], n = n, with_ties = FALSE)

  data |>
    filter(est_id %in% latest$est_id) |>
    mutate(est_id = as.character(est_id))
}
```

# Workforce

## Headcount

Number of people working in government on each date.

```{r headcount}
workforce$headcount |>
  plot_trend(group_col = "ref_date", y_col = "headcount", y_label = "Headcount")
```

## Headcount by establishment

The five establishments with the most staff on the latest date.

```{r headcount-by-est}
workforce$headcount_by_est |>
  top_establishments("headcount") |>
  plot_trend(group_col = "est_id", y_col = "headcount", y_label = "Headcount") +
  labs(color = "Establishment")
```

## Hires and separations

Hires are people active on a date but not on the previous one; separations
are people active on a date but not on the next one, including retirements.
Both are shown as a share of headcount.

```{r hires}
plot_movement(
  workforce$movement,
  movement_type = "hire",
  measurement_type = "rate"
)
```

```{r separations}
plot_movement(
  workforce$movement,
  movement_type = "separation",
  measurement_type = "rate"
)
```

## Moves between establishments

The ten most common moves between establishments, added up over all dates.

```{r transitions}
top_moves <- workforce$transitions |>
  summarise(moves = sum(transitions), .by = c(from, to)) |>
  slice_max(moves, n = 10, with_ties = FALSE)

if (nrow(top_moves) == 0) {
  cat("No moves between establishments were recorded.")
} else {
  knitr::kable(top_moves, col.names = c("From", "To", "Moves"))
}
```

# Wagebill

## Total wagebill

Total pay across all contracts on each date.

```{r wagebill}
wagebill$wagebill |>
  plot_trend(group_col = "ref_date", y_col = "wagebill", y_label = "Wagebill")
```

## Wagebill growth

Change in the wagebill since the previous date.

```{r wagebill-growth}
wagebill$wagebill |>
  filter(!is.na(wagebill_growth)) |>
  plot_trend(group_col = "ref_date", y_col = "wagebill_growth", y_label = "Growth") +
  scale_y_continuous(labels = scales::label_percent()) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50")
```

## Wagebill by establishment

The five establishments with the largest wagebill on the latest date.

```{r wagebill-by-est}
wagebill$wagebill_by_est |>
  top_establishments("wagebill") |>
  plot_trend(group_col = "est_id", y_col = "wagebill", y_label = "Wagebill") +
  labs(color = "Establishment")
```

## Average wage

Average pay per contract on each date.

```{r wage}
wagebill$wage |>
  plot_trend(group_col = "ref_date", y_col = "wage", y_label = "Average wage")
```

## Pay composition

Share of the wagebill paid through each pay component.

```{r composition}
wagebill$composition |>
  ggplot(aes(ref_date, share_wagebill, fill = component)) +
  geom_area() +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(x = "Time", y = "Share of wagebill", fill = "Component") +
  theme(legend.position = "bottom")
```

## Pay distribution

Share of contracts in each pay band, pooling all dates.

```{r distribution}
plot_histogram(wagebill$distribution)
```
